A Study of Common Bug Fix Patterns in Rust \thanks{Competing Interests: We have no competing interests and are funded by a Discovery Grant from Canada's Natural Science and Engineering Research Council.}

### 2023-05-18 results.tex
added actionability paragraphs to results section

We collected 60583 datapoints for general fix patterns using a server with a Xeon Gold 5120 processor (Q3-2017, 14-core 2.2 to 3.2GHz) and the data collection took around 26 hours to compute both $D_g$ and $D_b$. We used $\epsilon=0.0001$ and $Z=5$ as our DBSCAN parameters and that resulted in 577 clusters. After manual analysis, as described in~\ref{sec:manual_analysis_cluster_selection}, we ended up with 12 general patterns, which we describe in Table~\ref{table:general}. As shown in the table, we have separated the patterns based on the underlying program element (Category column). We have given each pattern a specific ID (ID column) and determined the number of instances manifesting that fix pattern (\# Instances column), as well as the number of projects in which the pattern was seen (\# Project column). Now we provide a detailed description for each of these patterns:

@@

Attributes in Rust enable developers to add metadata to program elements. Developers can add attributes to structs, struct fields, enum variants, match expression arms, etc. There are two types of attributes: Inner Attributes and Outer Attributes. Inner Attributes apply to the program element encapsulating them, while Outer Attributes apply to the program element coming after the attribute. \\

@@

\noindent\textbf{Description:} In this pattern, the attribute set of a struct field changes. As with \textit{\textbf{G.attr.struct}}, there are also three subtypes of this change pattern. Using these attributes, the developers can control different actions they wish to apply on struct fields, in addition to providing meta-information. For instance, in the code snippet provided above, the developers chose to make Serde (a serialization/deserialization tool for Rust) call a different function (\verb+is_empty+ instead of \verb+Option::is_none+) to determine whether to skip serializing the \verb+attributes+ field.

@@

\noindent\textbf{Description:} Adding \texttt{pub} to a struct field makes it possible to access that field from external modules. In the instances of this cluster, adding the \texttt{pub} access modifier happened in bug fixing changes, where the developer needed to access a field which had not been marked for public access. On the other hand, the public access might be revoked if the developer realizes there are no external accesses to the field, and that such accesses are not desirable, e.g. for encapsulation purposes.

@@

\noindent\textbf{Description:} Rust is a null-safe language, meaning that object references cannot take on null values. If a developer wants a variable to contain the equivalent of null, they would have to use the \texttt{Option} type. \texttt{Option} is essentially an \texttt{enum} with two variants, \texttt{Some}, which indicates that a variable has a value, and \texttt{None}, which indicates the absence of any value. In this change pattern, the developer changes the type of a struct field to an \texttt{Option} of that type, hence enabling them to store no value in that field.

For instance, in the Rust project swc-project/swc (a fast TypeScript/JavaScript compiler), the developer needed to modify the CSS parser source code to account for empty @media queries\footnote{\scriptsize \url{https://github.com/swc-project/swc/commit/75a14f98b7370226115ee24eec6eb8c802bd4837}}. As shown in the snippet above, they changed type \texttt{MediaQueryList} to \verb+Option<MediaQueryList>+, and modified the other parts of the source code accordingly.

@@

We collected 27143 datapoints for borrow-checker related fix patterns. We used $\epsilon=0.001$ and $Z=5$ as our DBSCAN parameters and that resulted in 102 clusters. After manual analysis, as described in~\ref{sec:manual_analysis_cluster_selection}, we ended up with 8 borrow-checker related fix patterns. Similar to the previous table, we describe BC-related patterns in Table~\ref{table:bc}. Now, we provide a detailed description of each of these patterns:

@@

@@

@@

>>>>

We collected 60583 datapoints for general fix patterns using a server with a Xeon Gold 5120 processor (Q3-2017, 14-core 2.2 to 3.2GHz) and the data collection took around 26 hours to compute both $D_g$ and $D_b$. We used $\epsilon=0.0001$ and $Z=5$ as our DBSCAN parameters and that resulted in 577 clusters. After manual analysis, as described in~\ref{sec:manual_analysis_cluster_selection}, we ended up with 12 general patterns, which we describe in Table~\ref{table:general}. As shown in the table, we have separated the patterns based on the underlying program element (Category column). We have given each pattern a specific ID (ID column) and determined the number of instances manifesting that fix pattern (\# Instances column), as well as the number of projects in which the pattern was seen (\# Project column). Since these patterns are the general Rust patterns, it is worth mentioning that similar patterns can also be observed in other programming languages. To make this more clear, we will highlight these similarities in each pattern's description. We will also discuss the actionability of some of the patterns. Moreover, irrespective of their actionability, the empirical frequency numbers associated with each pattern can serve as valuable guidance for program repair tools. These frequency numbers provide insights into the importance and prevalence of the bugs, enabling repair tools to prioritize and address them accordingly. Now we provide a detailed description for each of these patterns:

@@

Attributes in Rust enable developers to add metadata to program elements. Developers can add attributes to structs, struct fields, enum variants, match expression arms, etc. There are two types of attributes: Inner Attributes and Outer Attributes. Inner Attributes apply to the program element encapsulating them, while Outer Attributes apply to the program element coming after the attribute. The concept of attributes in Rust is similar to Java's annotations and Python's decorators. While each language may have its own syntax and specific use cases, the fundamental idea of attaching additional information to code elements is shared across these languages. \\

@@

\noindent\textbf{Description:} In this pattern, the attribute set of a struct field changes. As with \textit{\textbf{G.attr.struct}}, there are also three subtypes of this change pattern. Using these attributes, the developers can control different actions they wish to apply on struct fields, in addition to providing meta-information. For instance, in the code snippet provided above, the developers chose to make Serde (a serialization/deserialization tool for Rust) call a different function (\verb+is_empty+ instead of \verb+Option::is_none+) to determine whether to skip serializing the \verb+attributes+ field. \\

\noindent\textbf{Actionability:} Bug fix patterns related to attributes in Rust, such as adding or removing attributes can be valuable in designing program repair tools. For instance, by detecting and removing attributes that are no longer needed, program repair tools can optimize the codebase and eliminate unnecessary overhead. On the other hand, when adding an attribute, these patterns offer insights into how repair tools can manipulate attributes and modify other parts of the code accordingly to make it adhere to the user's specifications.

@@

\noindent\textbf{Description:} Adding \texttt{pub} to a struct field makes it possible to access that field from external modules. In the instances of this cluster, adding the \texttt{pub} access modifier happened in bug fixing changes, where the developer needed to access a field which had not been marked for public access. On the other hand, the public access might be revoked if the developer realizes there are no external accesses to the field, and that such accesses are not desirable, e.g. for encapsulation purposes. \\

\noindent\textbf{Actionability:} The patterns related to access modifiers offer an opportunity for repair tools to make sure that the reachability of program entities follows the user's specifications, enhancing security and maintainability. Additionally, the empirical numbers revealing the frequency of struct field modifications in Rust shed light on the significance of these bug fix patterns and emphasize the need for further exploration on how program repair tools can effectively modify code based on changes in struct fields to repair program behavior. These patterns provide valuable insights for program repair tools to automatically analyze and modify the codebase, ensuring it aligns with the desired behavior.

@@

\noindent\textbf{Description:} Rust is a null-safe language, meaning that object references cannot take on null values. If a developer wants a variable to contain the equivalent of null, they would have to use the \texttt{Option} type. \texttt{Option} is essentially an \texttt{enum} with two variants, \texttt{Some}, which indicates that a variable has a value, and \texttt{None}, which indicates the absence of any value. In this change pattern, the developer changes the type of a struct field to an \texttt{Option} of that type, hence enabling them to store no value in that field. 
For instance, in the Rust project swc-project/swc (a fast TypeScript/JavaScript compiler), the developer needed to modify the CSS parser source code to account for empty @media queries\footnote{\scriptsize \url{https://github.com/swc-project/swc/commit/75a14f98b7370226115ee24eec6eb8c802bd4837}}. As shown in the snippet above, they changed type \texttt{MediaQueryList} to \verb+Option<MediaQueryList>+, and modified the other parts of the source code accordingly. \\

\noindent\textbf{Actionability:} When the user's specification requires a variable to be able to assume \texttt{None} values, these patterns indicate the need to modify the program accordingly. A program repair tool can leverage these patterns to automatically update the variable's type to \texttt{Option<T>}, ensuring that it can now hold \texttt{None} as a valid value. Furthermore, such a tool should also systematically update other occurrences of that variable throughout the program to account for the change and ensure program correctness.

@@

We collected 27143 datapoints for borrow-checker related fix patterns. We used $\epsilon=0.001$ and $Z=5$ as our DBSCAN parameters and that resulted in 102 clusters. After manual analysis, as described in~\ref{sec:manual_analysis_cluster_selection}, we ended up with 8 borrow-checker related fix patterns. Similar to the previous table, we describe BC-related patterns in Table~\ref{table:bc}. It is important to note that due to the unique nature of the Borrow-Checker as an exclusive component of Rust, these patterns may not have direct counterparts or equivalents in other languages. Now we provide a detailed description of each of these patterns:

@@

\noindent\textbf{Actionability:} A program repair tool can efficiently implement this modification by examining the necessity of the clone operation and considering the user's specifications regarding variable cloning. If the clone is deemed unnecessary, the tool can effortlessly replace it with a reference to reduce the overhead of unnecessary copying, as discussed in the example above. \\

@@

\noindent\textbf{Actionability:} This pattern often occurs as part of a larger change, making it challenging to create a generalized repair tool that covers all possible scenarios. Typically, when a reference is passed instead of transferring ownership, there is subsequent usage of the variable after the function call. Although creating a comprehensive repair tool for such cases is complex, empirical results underscore the significance and prevalence of this fix, warranting further investigation. \\

@@

\noindent\textbf{Actionability:} A repair tool can utilize this pattern to meet the user's specifications by introducing mutability to the target variable and modifying other parts of the program accordingly. For instance, if the target variable has associated \texttt{impl} blocks, the change may involve modifying method annotations to ensure that the calling object is mutable, aligning with the desired specifications. \\
